Approximation as Differentiation

### 2025-07-03 main/first-order.tex
Fix enumerated list inside example environment.
Simplify AD section a bit, add something about reverse-mode in Fam(FinVect^op).

% which can then be composed with $\partial f: \partial X \to T(f)(\partial Y)$.

@@

As per the CHAD work, automatic differentiation for first-order programs can also be recovered in this setup,
using the total category $\Fam(\FinVect)$ instead of $\Fam(\LatGal)$. First we recall the chain rule for
forward derivatives $\pushf{f}_x: \RR^m \linearto \RR^n$ and backward derivatives $\pullf{f}_x: \RR^n
\linearto \RR^m$, which says that derivatives respect composition~\cite{spivak65}.

@@

\subsubsection{Forward-Mode Automatic Differentiation}

The basic idea of (forward-mode) automatic differentiation, as first described by \citet{linnainmaa76}, was to
recognise that one could efficiently compute the pushforward map $\pushf{f}: \RR^m \to (\RR^m \linearto
\RR^n)$ of a differentiable function $f: \RR^m \to \RR^n$ alongside the computation of $f$ itself, by defining
a single map $\tangents(f) = x \mapsto (f(x), \pushf{f}_x): \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$.
Defining $f_1 = \pi_1 \comp f: \RR^m \to \RR^n$ and $f_2 = \pi_2 \comp f: \RR^m \to (\RR^m \linearto \RR^n)$,
composition of such maps can be expressed as:
\begin{align*}
(g \comp f)_1(x) &= f_1(g_1(x)) \\
(g \comp f)_2(x) &= g_2(f_1(x)) \comp f_2(x)
\end{align*}

We can verify that $\tangents$ is functorial, regardless of whether the linear maps $\RR^m \linearto \RR^n$
happen to be derivatives. But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and
similarly for $g$ then the derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is
the derivative of $(g \comp f)_1$ at $x$. Thus this setup provides an algorithmic method for computing the
forward derivative of a composite function, as long as all atomic operations provide derivatives, without have
to recompute the function itself.

@@

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the family of tangent spaces which is constantly $\RR^n$, and each map $f:
\RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as the pair $(f_1, f_2)$. Then the composition $(g, \partial
g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp \partial f)$ again corresponds exactly to
the forward chain rule, since $(\reindex{\partial g}{f} \comp \partial f)_x = {\partial g}_{f(x)} \comp
{\partial f}_x$.

% with the reindexing along $f$ recomputing $f(x)$ in order to select the appropriate pushforward.

\subsubsection{Reverse-Mode Automatic Differentiation}

\todo{brief summary}

>>>>

@@

As per the CHAD work, automatic differentiation (AD) for first-order programs can also be recovered in this
setup, using the total categories $\Fam(\FinVect)$ and $\Fam(\FinVect^\op)$ instead of $\Fam(\LatGal)$. First
we recall the chain rule for forward derivatives $\pushf{f}_x: \RR^m \linearto \RR^n$ and backward derivatives
$\pullf{f}_x: \RR^n \linearto \RR^m$, which says that derivatives respect composition~\cite{spivak65}.

@@

\subsubsection{Automatic Differentiation}

The basic idea of (forward-mode) AD, as first described by \citet{linnainmaa76}, was to recognise that one
could efficiently compute the pushforward map $\pushf{f}: \RR^m \to (\RR^m \linearto \RR^n)$ of a
differentiable function $f: \RR^m \to \RR^n$ alongside the computation of $f$ itself, by defining a single map
$\tangents(f) = x \mapsto (f(x), \pushf{f}_x): \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$. Writing $f_1$
for $\pi_1 \comp f$ and $f_2$ for $\pi_2 \comp f$, composition of such maps can be expressed as:
\begin{align*}
(g \comp f)_1(x) &= f_1(g_1(x)) \\
(g \comp f)_2(x) &= g_2(f_1(x)) \comp f_2(x)
\end{align*}

We can verify that $\tangents$ is functorial, regardless of whether the linear maps happen to be derivatives.
But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and similarly for $g$ then the
derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is the derivative of $(g \comp
f)_1$ at $x$. Thus this setup provides an algorithmic method for computing the forward derivative of a
composite function, as long as all atomic operations provide derivatives, without have to recompute the
function itself.

@@

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the family of tangent spaces which is constantly $\RR^n$, and each map $f:
\RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as a pair of maps $(f_1, f_2)$. Then the composition $(g,
\partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp \partial f)$ again corresponds
exactly to the forward chain rule, since $(\reindex{\partial g}{f} \comp \partial f)_x = {\partial g}_{f(x)}
\comp {\partial f}_x$. Reverse-mode AD can be interpreted similarly in $\Fam(\FinVect^\op)$, reading the
single map $g: \RR^m \to \RR^n \times (\RR^n \linearto \RR^m)$ as the pair of maps $(g_1, g_2)$.

### 2025-07-03 main/language.tex
Tweak.
Simplify AD section a bit, add something about reverse-mode in Fam(FinVect^op).

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(\coprod, 0)$ and exponentials
$\internalHom{X}{Y}$, with evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$,
plus the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$;
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}_{\PrimTy}$;
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}

\figrefTwo{semantics:types}{semantics:contexts} give the interpretations of types $\tau$ and contexts $\Gamma$
as objects of $\Sem$. \figref{semantics:terms} gives the interpretation of terms $\Gamma \vdash t: \tau$ as
morphisms $\sem{\Gamma} \to \sem{\tau}$.

>>>>

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(\coprod, 0)$ and exponentials
$\internalHom{X}{Y}$, with evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$,
plus the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}_{\PrimTy}$
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}

\figref{semantics:types} gives the interpretations of types $\tau$ and contexts $\Gamma$ as objects of $\Sem$.
\figref{semantics:terms} gives the interpretation of terms $\Gamma \vdash t: \tau$ as morphisms $\sem{\Gamma}
\to \sem{\tau}$.
